Revealing transient strain in geodetic data with Gaussian process regression

### 2017-07-14 RevisedSubmission/hines_hetland_2017_gji.tex
started making edits

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive-definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer transient strain resulting from SSEs in Cascadia. Our results demonstrate that GPR is an effective tool for detecting SSEs and revealing how strain evolves during SSEs.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot\varepsilon(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacements independently and ignore any potential covariance. Hence, we drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into positive definite spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

@@

The prior for transient displacements is then conditioned with $\mitbf{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \mitbf{d}_*$. For now, we will assume that appropriate covariance functions and corresponding hyperparameters for $X$, $T$, and $C_\eta$ have already been chosen. In Section \ref{sec:NoiseModel} and \ref{sec:SignalModel}, we discuss how the covariance functions are chosen for our application to GNSS data from Cascadia. If $\kappa$ is kept finite then, following \citet{Rasmussen2006}, we find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(p) = C_u(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}\mitbf{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(p,p') = C_u(p,p') - C_u(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}C_u(\mitbf{P},p').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We use the partitioned matrix inversion identity \citep[e.g.,][]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\mitbf{P}) & \mitbf{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \mitbf{\Sigma} & \mitbf{G} \\
                         \mitbf{G}^T  & -\kappa^{-2} \mitbf{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \mitbf{d}_* \\
                         \mitbf{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\mitbf{P}) & \mitbf{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \mitbf{\Sigma} & \mitbf{G} \\
                          \mitbf{G}^T  & -\kappa^{-2} \mitbf{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\mitbf{P},p') \\
                          \mitbf{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\mitbf{P}) & \mitbf{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \mitbf{\Sigma} & \mitbf{G} \\
                         \mitbf{G}^T  & \mitbf{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \mitbf{d}_* \\
                         \mitbf{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\mitbf{P}) & \mitbf{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \mitbf{\Sigma} & \mitbf{G} \\
                          \mitbf{G}^T  & \mitbf{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\mitbf{P},p') \\
                          \mitbf{0} \\
                          \end{array}\right].
\end{equation}

We use eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to find the posterior easting and northing components of transient displacements. Using $\hat{u}_i$ to denote the posterior transient displacements along direction $i$ and $x_i$ to represent the components of $\vec{x}$, we can write the components of $\dot\varepsilon$ as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The transient strain rate components are Gaussian processes with mean functions
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance functions
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}

>>>>

Here we propose using Gaussian process regression (GPR) \citep{Rasmussen2006} to estimate transient strain from GNSS data. GPR is a Bayesian, non-parametric method for inferring a continuous signal from scattered data. Since GNSS stations are irregularly spaced and observation times may differ between stations, GPR is an ideal tool for synthesizing GNSS data into a spatially and temporally continuous representation of surface deformation. GPR is closely related to kriging \citep{Cressie1992} and least squares collocation \citep{Moritz1978}. The latter has been used by \citet{Kato1998} and \citet{El-Fiky1999} to estimate secular strain rates from GNSS data. GPR is Bayesian in that we describe our prior understanding of the geophysical signal with a Gaussian process. A Gaussian process is a normally distributed stochastic process that is fully defined in terms of a mean function and a positive definite covariance function. For example, Brownian motion, $B(t)$, is a well known Gaussian process in $\mathbb{R}^1$ which has zero mean and covariance function $\mathrm{cov}(B(t),B(t')) = \min(t,t')$, where $t,t' \ge 0$. If no prior information is available for the geophysical signal, then maximum likelihood methods can be used to choose a prior Gaussian process that is most consistent with the observations.  We incorporate GNSS observations with the prior to form a posterior estimate of transient strain.  The posterior transient strain is also a Gaussian process, and we can use its distribution to confidently discern geophysical signal from noise. We use GPR to infer transient strain resulting from SSEs in Cascadia. Our results demonstrate that GPR is an effective tool for detecting SSEs and revealing how strain evolves during SSEs.

\section{Estimating Transient Strain Rates}\label{sec:Method}
We seek a spatially and temporally dependent estimate of transient crustal strain rates. We consider transient strain rates to be any deviation from secular strain rates, and our attention is limited to horizontal strain rates in this study. We denote transient crustal strain rates as $\dot\varepsilon(p)$, where $p$ represents the ordered pair $(\vec{x},t)$, $\vec{x}$ are spatial coordinates in $\mathbb{R}^2$, and $t$ is time. We determine $\dot\varepsilon(p)$ by spatially and temporally differentiating estimates of transient displacements, $\vec{u}(p)$. We make a prior assumption that each component of $\vec{u}$ is a Gaussian process,
\begin{equation}\label{eq:TransientDeformation}
u_i(p) \sim \mathcal{N}\left(0,C_{u_i}\right),
\end{equation}
where $C_{u_i}(p,p')$ is a covariance function indicating how we expect $u_i(p)$ to covary with $u_i(p')$. For simplicity, we treat each component of displacements independently and ignore any potential covariance between them. Hence, we drop the component subscripts with the understanding that the same analysis is being repeated to estimate the easting and northing components of $\vec{u}$. We assume that $C_u$ can be separated into spatial and temporal functions as 
\begin{equation}\label{eq:TransientCovariance}
C_{u}\left((\vec{x},t),(\vec{x}',t')\right) = X(\vec{x},\vec{x}')T(t,t').
\end{equation}  
As long as the functions $X$ and $T$ are positive definite, $C_u$ is guaranteed to also be positive definite and thus a valid covariance function \citep[sec. 4.2.4]{Rasmussen2006}. The appropriate choice for $X$ and $T$ may vary depending on the geophysical signal we are trying to describe, and we discuss this matter in Section \ref{sec:SignalModel}.

@@

The prior for transient displacements is then conditioned with $\mitbf{d}_*$ to form a posterior estimate of transient displacements, $\hat{u} = u | \mitbf{d}_*$. For now, we will assume that appropriate covariance functions and corresponding hyperparameters for $C_u$ and $C_\eta$ have already been chosen. In Section \ref{sec:NoiseModel} and \ref{sec:SignalModel}, we discuss how the covariance functions are chosen for our application to GNSS data from Cascadia. If $\kappa$ is kept finite then, from \citet[sec. 2.2]{Rasmussen2006}, we find that $\hat{u}$ is distributed as $\mathcal{N}(\mu_{\hat{u}},C_{\hat{u}})$, where
\begin{equation}\label{eq:PosteriorMean}
\mu_{\hat{u}}(p) = C_u(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}\mitbf{d}_*
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov}
C_{\hat{u}}(p,p') = C_u(p,p') - C_u(p,\mitbf{P})\left(\mitbf{\Sigma} + \kappa^2\mitbf{G}\mitbf{G}^T\right)^{-1}C_u(\mitbf{P},p').
\end{equation}
However, we are interested in the limit as $\kappa \to \infty$, and the form for eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) is not suitable for evaluating this limit. We use a partitioned matrix inversion identity \citep[sec. 2.7.4]{Press2007} to rewrite eq. (\ref{eq:PosteriorMean}) and eq. (\ref{eq:PosteriorCov}) as
 \begin{equation}\label{eq:PosteriorMean2}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\mitbf{P}) & \mitbf{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \mitbf{\Sigma} & \mitbf{G} \\
                         \mitbf{G}^T  & -\kappa^{-2} \mitbf{I} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \mitbf{d}_* \\
                         \mitbf{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov2}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\mitbf{P}) & \mitbf{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \mitbf{\Sigma} & \mitbf{G} \\
                          \mitbf{G}^T  & -\kappa^{-2} \mitbf{I} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\mitbf{P},p') \\
                          \mitbf{0} \\
                          \end{array}\right].
\end{equation}
Taking the limit as $\kappa \to \infty$, we get the solution for the mean and covariance of $\hat{u}$,
 \begin{equation}\label{eq:PosteriorMean3}
\mu_{\hat{u}}(p) = \left[\begin{array}{cc}
                         C_u(p,\mitbf{P}) & \mitbf{0} \\
                         \end{array}\right]
                   \left[\begin{array}{cc}
                         \mitbf{\Sigma} & \mitbf{G} \\
                         \mitbf{G}^T  & \mitbf{0} \\
                         \end{array}\right]^{-1}
                   \left[\begin{array}{c}
                         \mitbf{d}_* \\
                         \mitbf{0} \\
                         \end{array}\right]
\end{equation}    
and
\begin{equation}\label{eq:PosteriorCov3}
C_{\hat{u}}(p,p') = C_u(p,p') - 
                    \left[\begin{array}{cc}
                          C_u(p,\mitbf{P}) & \mitbf{0} \\
                          \end{array}\right]
                    \left[\begin{array}{cc}
                          \mitbf{\Sigma} & \mitbf{G} \\
                          \mitbf{G}^T  & \mitbf{0} \\
                          \end{array}\right]^{-1}
                    \left[\begin{array}{c}
                          C_u(\mitbf{P},p') \\
                          \mitbf{0} \\
                          \end{array}\right].
\end{equation}

We use eq. (\ref{eq:PosteriorMean3}) and (\ref{eq:PosteriorCov3}) to find the posterior easting and northing components of transient displacements. Using $\hat{u}_i$ to denote the posterior transient displacements along direction $i$ and $x_i$ to represent the components of $\vec{x}$, we can write the components of $\dot\varepsilon$ as 
\begin{equation}\label{eq:StrainRate}
\dot\varepsilon_{ij}(p) = \frac{1}{2} \frac{\partial}{\partial t} \left(
                                        \frac{\partial \hat{u}_i(p)}{\partial x_j} +  
                                        \frac{\partial \hat{u}_j(p)}{\partial x_i}\right).
\end{equation}
The transient strain rate components are stochastic processes whose distributions can be interpreted as the distributions of samples of $\hat{u}$ that have been differentiated by the operator from eq. (\ref{eq:StrainRate}). Since differentiation is a linear operation, the transient strain rate components are Gaussian processes. From \citet[sec. 10.2]{Papoulis1991}, we find that the components of $\dot{\varepsilon}$ have mean functions
\begin{equation}\label{eq:StrainMean}
\mu_{\dot\varepsilon_{ij}}(p) = \frac{1}{2}\frac{\partial}{\partial t}\left(
                                  \frac{\partial \mu_{\hat{u}_i}(p)}{\partial x_j} + 
                                  \frac{\partial \mu_{\hat{u}_j}(p)}{\partial x_i} \right)
\end{equation} 
and covariance functions
\begin{equation}\label{eq:StrainCov}
C_{\dot\varepsilon_{ij}}(p,p') = \frac{1}{4} \frac{\partial^2}{\partial t \, \partial t'}\left(
                                   \frac{\partial^2 C_{\hat{u}_i}(p,p')}{\partial x_j \, \partial x'_j} + 
                                   \frac{\partial^2 C_{\hat{u}_j}(p,p')}{\partial x_i \, \partial x'_i} \right).
\end{equation}
